---
title: "Multilevel Bayesian Regression model"
author: "Axel Arango"
date: "2022-11-03"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
This is the code to perform the Multilevel Baysian Regression Model(s) to test the prevalence of the relationship between dispersal ability evolutionary rate and speciation rate across the Emberizoidea superfamily:

First, we will load the necessary packages: 

```{r, eval= T, include=T,warning=FALSE}
library(tidyverse)
library(brms)
library(cmdstanr)
```

Then, the necessary data.

In this case we are only going to use the branch lengths obtained using the Cor-STRATES framework, and separated them by families.

```{r, eval=T,include=T}
dat<-read_csv("edgmeasures.csv")
dat2<-data.frame(family=dat$family,loglambda=log(dat$edge_lambda),logbeta=log(dat$edge_beta))
head(dat2)# this file contains the log transformed values for the speciation and phenotypic evolutionary rates per branch per taxonomic family of Emberizoidea
```

Once we have loaded the data we can start to fit the Bayesian multilevel models.

First we are going to set up the initial conditions.

```{r, eval=T,include=T}
iters = 5000 #number of iterations
burns = iters / 4 # 25% as burnings
nCores = 2 #number of cores for our computer to use
nChains = 4 #number of Markov chains
```

With the initial values set up, we can fit our Bayesian model.
In this model we are going to evaluate the variation of the speciation within each family.

```{r, eval=T,include=T,message=FALSE,warning=FALSE}
fit_All_within<- brm(logbeta~loglambda  + (1 + loglambda| family), 
                      data = dat2, 
                      chains = nChains, 
                      iter = iters, 
                      cores = nCores,
                      backend = "cmdstanr")
summary(fit_All_within)
```

Here we found that the estimated standard deviation (sd) of the relationship between dispersal ability evolutionary rate and speciation rate varied substantially (Mean sd = 0.69 [0.45:1.08; 95% credible intervals])

Let's explore further to estimate the effect of the association between the rates of trait evolution and speciation for each family

First, we will extract the intercepts of species varying within family using several credible intervals 

```{r, eval=T,include=T,message=FALSE,warning=FALSE}
fam_ALL_within <- ranef(fit_All_within, probs = c(0.025, 0.05, 0.11, 0.89, 0.95, 0.975))$family
```

Then we calculate the prevalence of such family intercepts, to facilitate visualization we are going to load first a custom function and establish the colors for the intercept deviations

```{r, eval=T,include=T}
load("theme_nice.R")
barCOLS <- c(scales::alpha("red", 0.7), 
             scales::alpha("darkblue", 0.7), 
             scales::alpha("darkgray", 0.7))
```

Now we evaluate the effect on the association per family

```{r, eval=TRUE,include=TRUE, warning=FALSE}
p_family <- fam_ALL_within[, , "Intercept"] %>%  
  data.frame() %>% 
  rownames_to_column() %>%
  rename(family = rowname) %>% 
  as_tibble() %>% 
  mutate(direction = ifelse(Q11 > 0, "Positive", 
                            ifelse(Q89 < 0, "Negative", "Null"))) %>% 
  arrange(family) %>% 
  ggplot(aes(x = family, y = Estimate, color = direction)) +  
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5), 
                  fatten = 7, linewidth = 1.5, alpha = 0.4) + 
  geom_pointrange(aes(ymin = Q11, ymax = Q89), 
                  fatten = 10, linewidth = 2.5, alpha = 0.7) + 
  #coord_cartesian(ylim = c(-1, 1.5)) + 
  geom_hline(yintercept = 0, linewidth = 1, colour = "darkgray", linetype = "dashed") + 
  scale_colour_manual(values = c("Positive" = barCOLS[2], 
                                 "Negative" = barCOLS[1], 
                                 "Null" = barCOLS[3])
  ) + 
  coord_flip() + 
  labs(x = NULL, 
       y = "Family-level effect - SD") + 
  theme_nice() + 
  theme(
    legend.position = "none", 
    axis.text.x = element_text(vjust = 0.5, hjust = 0.7), 
    axis.text.y = element_text(face = "italic"),
    axis.title = element_text(size = 18), 
    axis.text = element_text(size = 15, colour = "black")
  ) 

plot(p_family)
```

Results show that the families Thraupidae and Passerellidae have a strong negative effect on the association between the rates of trait evolution and speciation. In contrast, Emberizidae and Cardinallidae show a positive effect—although the 95% credible intervals overlap with zero 